Fast and scalable Gaussian process modeling of stellar variability \vspace{-3\baselineskip}

### 2016-10-05 paper/ms.tex
Added some intro, refs

@@

\begin{enumerate}
\item Non-parametric modeling of time series with correlated noise using Gaussian processes
\item Assumptions:\\
  a). Stationary noise (although this can be broken).\\
  b). Gaussian
\item Problem: can't handle large datasets - e.g. entire Kepler long-cadence; Spitzer light
   curves; Solar curves; short-cadence.
\item Solution(s):\\
  a). HODLR \citep{Ambikasaran2013,Ambikasaran2016}\\
  b). CARMA (Lorentzian power spectra, but with limitations) \citep{Kelly2014}\\
  c). Tri-diagonal (Press \& Rybicki) \citep{1995PhRvL..74.1060R}
\item Problem: HODLR is tricky to use; CARMA requires certain relations between
  coefficients of Lorentzians (right?); P\&R only works for up to 2 components + white noise
\item Solution: Ambikasaran \citep{Ambikasaran2015} shows how sum of exponential kernels can be solved in order
   $O(N*p^2)$, where p is number of Lorentzians.
\item Generalized Ambikasarn for complex coefficients (Hermitian).
\item Performance:\\
  a). Show $O(N*p^2)$ scaling\\
  b). Show comparison to HODLR.\\
  c). Show comparison to wavelet (ala Carter \& Winn).\\
  d). Compare with CARMA Kallman filter(?).
\item Various methods/formulae:\\
  a). Likelihood computation.\\
  b). Generating GP with particular covariance [yet to solve]. [ ]\\
  c). Inferring particular components.\\
  d). Computing derivatives of GPs (as used in RV methods).\\
  e). Forecasting/interpolating.\\
  f). Inferring confidence intervals.\\
  g). Derivative of likelihood function with respect to kernel parameters.\\
  h). Non-stationary GP (can the coefficients vary with time?).\\
  i). Multi-band (each element of extended matrix becomes a matrix).

\item Generalization to non-stationary noise.
\item Example application: TYC 3559\\
  a). Entire long-cadence 17-quarter Kepler dataset (~65k data points).\\
  b). Analysis of solar data.

\item Possible future applications:\\
  a). Inference of asteroseismic variability - ala Andrew Gordon Wilson \citep{2013arXiv1302.4245W}, but
      with Lorentzians, not Gaussians.\\
  b). Multi-waveband GPs.\\
  c). RV interpolation.\\
  d). Non-parametric phase functions.\\
  e). Better calibration of the flicker log(g) relation.\\
  f). Measurement of rotational periods - gyrochronology.\\
  g). Doppler shifts in oscillating EBs.

>>>>

Stars are variable, for better or worse:  their variability provides information about the properties
of stars, but also results in noise that must be overcome in detecting and characterizing their
planetary systems.  When analyzing astronomical times series of stars, then, there are generally two 
goals:  1). to analyze the frequency spectrum of stellar variability to beter understand stellar
properties and evolution, and/or 2). to account for correlated noise when modeling the stellar
flux variability and spectral variations.

A promising technique for addressing both of these goals is the Gaussian Process formalism 
\citep{Rasmussen, Gibson}.  The basis of this method is to assume that stellar variability
is random, but that the random variability is correlated in time, and the nature of the
correlations is completely specified by a correlation matrix.  In its simplest form, the 
correlation matrix is modeled with a time-independent autocorrelation function (although 
the time-independence can be relaxed), including a white-noise component that may vary
with time, and the autocorrelation function is described by a functional form that may be 
simply parameterized by a relatively small number of parameters, sometimes referred to as 
`hyper-parameters.'  The goal of spectral analysis is to measure the parameters which
describe the autocorrelation function, thus inferring properties of the stellar variabilty
which may be used to characterize properties of the star, such as the rotation period,
the asteroseismic frequencies, or the `flicker' \citep{Aerts2010,Noyes1984,Bastien2013}.

@@

\begin{enumerate}
\item Non-parametric modeling of time series with correlated noise using Gaussian processes
\item Assumptions:\\
  a). Stationary noise (although this can be broken).\\
  b). Gaussian
\item Problem: can't handle large datasets - e.g. entire Kepler long-cadence; Spitzer light
   curves; Solar curves; short-cadence.
\item Solution(s):\\
  a). HODLR \citep{Ambikasaran2013,Ambikasaran2016}\\
  b). CARMA (Lorentzian power spectra, but with limitations) \citep{Kelly2014}\\
  c). Tri-diagonal (Press \& Rybicki) \citep{1995PhRvL..74.1060R}
  d). KISS-GP \citep{WilsonNickisch2015} %Kernel Interpolation for Scalable Structured Gaussian Processes (KISS-GP) Andrew Wilson, Hannes Nickisch Proceedings of The 32nd International Conference on Machine Learning, pp. 1775-1784, 2015
\item Problem: HODLR is tricky to use; CARMA requires certain relations between
  coefficients of Lorentzians (right?); P\&R only works for up to 2 components + white noise
\item Solution: Ambikasaran \citep{Ambikasaran2015} shows how sum of exponential kernels can be solved in order
   $O(N*p^2)$, where p is number of Lorentzians.
  - Sum of Lorentzians is a good description of stellar variability:
     - granulation has expoential (or sums of exponential) noise
     - asteroseismic obeys sums of Lorentzians
     - periodic variability due to spots/plages/etc. can also be
       approximated by Lorentzian(s)
\item Generalized Ambikasarn for complex coefficients (Hermitian).
\item Performance:\\
  a). Show $O(N*p^2)$ scaling\\
  b). Show comparison to HODLR.\\
  c). Show comparison to wavelet (ala Carter \& Winn).\\
  d). Compare with CARMA Kallman filter(?).
\item Various methods/formulae:\\
  a). Likelihood computation.\\
  b). Generating GP with particular covariance [yet to solve]. [ ]\\
  c). Inferring particular components.\\
  d). Computing derivatives of GPs (as used in RV methods).\\
  e). Forecasting/interpolating.\\
  f). Inferring confidence intervals.\\
  g). Derivative of likelihood function with respect to kernel parameters.\\
  h). Non-stationary GP (can the coefficients vary with time?).\\
  i). Multi-band (each element of extended matrix becomes a matrix).
  j). Approximating various kernels: Matern, exp(sin^2), Gaussian, etc.
\item Generalization to non-stationary noise.
\item Example application: TYC 3559\\
  a). Entire long-cadence 17-quarter Kepler dataset (~65k data points).\\
  b). Analysis of solar data.
  c). CO2 Mauna Kea
\item Possible future applications:\\
  a). Inference of asteroseismic variability - ala Andrew Gordon Wilson \citep{2013arXiv1302.4245W}, but
      with Lorentzians, not Gaussians.\\
  b). Multi-waveband GPs.\\
  c). RV interpolation.\\
  d). Non-parametric phase functions.\\
  e). Better calibration of the flicker log(g) relation.\\
  f). Measurement of rotational periods - gyrochronology.\\
  g). Doppler shifts in oscillating EBs.
